---
title: "get_test_set"
author: "Maricar L. Rabonza"
date: "6/25/2021"
output: html_document
---

This file generates a train and test set from Dataset 1 and 2.
The test set is randomly selected from Dataset 1 only with 20-80 test train ratio.
All of Dataset 2 goes to the training set.

Raw data comes from `Data` folder.
Output files go to `Train and Test` folder.



```{r setup, include=FALSE}
#Working libraries
library(here) # sets location of .RProj as working directory
library(knitr) # useful for many rmarkdown layout functions
library(raster) # handling rasters
library(dplyr) # data wrangling
library(ggplot2) # the ultimate plotter
library(geosphere)
library(sp) # raster plotting
library(gstat) # helps with geostatistics and kriging
library(plotly) # another nice plotter
library(automap) # gives out pretty variogram plots
library(DescTools) # reverse dataframe order
library(GGally)
library(PerformanceAnalytics)
library(ggExtra)
library(gridExtra) # gridarrange
library(caret) # k-fold cross-validation
library(lemon) # grid_arrange_shared_legend
library(tidyverse)
library(rgeos) # for minimum distance
library(viridis) #color palette
library(ggpubr)
#library(cluster)
```

## Read data

```{r }
#-------------------
# READ ALL OBSERVED DATA
#-------------------

# Specify folder locations
kelud.df <- data.frame(easting=644183.3,northing=9123213.6) # input coord of volcano# Dataset 1
ds1_loc = here("Data", "datasets_raw", "Kelud_deposit_after2-3days_CVGHM_81points.csv") # Dataset 2
ds2_loc = here("Data", "datasets_raw", "Kelud_deposit_after3weeks_Maeno_50points.csv") # Dataset 2 folder

# Read dataset 1
allpts <- read.csv(ds1_loc, head=TRUE, sep=",")

# Read dataset 2 
allpts_ds2 <- read.csv(ds2_loc, head=TRUE, sep=",")
colnames(allpts_ds2) <- c('Easting', 'Northing', 'Load')

```

## Remove outlier in Dataset 1

```{r }

# Tag outlier: 629785, 9122495 (Load = 7)
allpts$Outlier <- "No"; allpts$Outlier[allpts$Easting == 629785] <- "Yes"

# Plot map of all points showing outlier as red
ggplot() +
    geom_point(data=kelud.df, aes(x=easting, y=northing), shape=24, bg="red", size=3) +
    geom_point(data = allpts, aes(x = Easting, y = Northing, size=Load), alpha = 0.3) +
    scale_size_continuous(name=bquote(atop(Tephra~load~phantom(), (kg/~m^2)))) + 
    geom_point(data=allpts[allpts$Outlier == "Yes", ], aes(x=Easting, y=Northing, size = Load),  color ="red", alpha = 0.3) +
    ggtitle("Dataset 1 (CVGHM)") +
    labs(x = "Easting (m)",
         y = "Northing (m)") +
    xlim(580000, 670000) +
    ylim(9102400, 9161100) +
    theme_bw()+
    theme(legend.position = c(.88, .5), legend.key = element_rect(fill = "white", colour = "black")) +
    cowplot::background_grid(major = "none", minor = "none") +
    coord_fixed(xlim = c(580000, 680000) , ylim = c(9102400, 9162400)) # zoom to map coords

# Remove points tagged as outlier
allpts <- allpts[allpts$Outlier == "No", ]
nrow(allpts) # show number of points once outlier/s are removed

# Remove tag column
allpts <- allpts[,-4]


```

## Histograms of data

```{r }
mean_load <- mean(allpts$Load)
ggplot(data=allpts, aes(x=Load)) +
    geom_histogram(binwidth = 5) + 
    ggtitle("Dataset 1 loads")  +
    geom_vline(xintercept=mean_load)+
    theme_bw()

mean_load_ds2 <- mean(allpts_ds2$Load)
ggplot(data=allpts_ds2, aes(x=Load)) +
    geom_histogram(binwidth = 5) + 
    ggtitle("Dataset 2 loads")  +
    geom_vline(xintercept=mean_load_ds2)+
    theme_bw()

# Remove very large values in Dataset 2 that produce very large errors
# Easting Northing Load 
# 645089.9	9126336	    3.50e+02	
# 641221.3	9121778	    2.10e+02	
# 647377.0	9126551	    2.24e+02

# Tag outlier: 629785, 9122495 (Load = 7)
allpts_ds2$Outlier <- "No"; 
allpts_ds2$Outlier[allpts_ds2$Load ==3.50e+02] <- "Yes"
allpts_ds2$Outlier[allpts_ds2$Load ==2.24e+02] <- "Yes"
allpts_ds2$Outlier[allpts_ds2$Load ==2.10e+02] <- "Yes"
# Remove points tagged as outlier
allpts_ds2 <- allpts_ds2[allpts_ds2$Outlier == "No", ]
nrow(allpts_ds2) # show number of points once outlier/s are removed
allpts_ds2 <- allpts_ds2[,-4] # Remove tag column

#Histogram after removing outlier
mean_load_ds2 <- mean(allpts_ds2$Load)
ggplot(data=allpts_ds2, aes(x=Load)) +
    geom_histogram(binwidth = 5) + 
    ggtitle("Dataset 2 loads - 3 large points removed")  +
    #xlim(c(-2,200)) +
    geom_vline(xintercept=mean_load_ds2)+
    theme_bw()

```


## Select random test points from Dataset 1

```{r}

# Set seed
set.seed(3)

# Set number of test points (m)
m <- 16 # with a 20-80 test train ratio
id <- sample(nrow(allpts), m) # randomly sample m points in the subset
test_set <- allpts[id,]
train_set <- anti_join(allpts, test_set, by = c("Easting" = "Easting")) # Set points not in test set as train points

# Tag training and test points
train_set$Set <- "Training"
test_set$Set <- "Test"
obs <- rbind(train_set, test_set)
obs$Set <- factor(obs$Set, levels = c("Training", "Test"))
```


## Check load distribution of test points
```{r}
hist_test  <- ggplot(data=test_set, aes(x=Load)) + geom_histogram(binwidth = 5) + ggtitle("Test set load") + 
    #xlim(c(0,250)) + 
    ylim(c(0,22)) +
    theme_bw()

hist_train  <-ggplot(data=train_set, aes(x=Load)) + geom_histogram(binwidth = 5) + ggtitle("Train set load") + 
    #xlim(c(0,250)) + 
    ylim(c(0,22)) +
    theme_bw()

hist_ds2  <-ggplot(data=allpts_ds2, aes(x=Load)) + geom_histogram(binwidth = 5) + ggtitle("Dataset 2 load") + 
    #xlim(c(0,250)) + 
    ylim(c(0,22)) +
    theme_bw()

ggarrange(hist_train, hist_test, hist_ds2,
                    ncol = 1, nrow = 3)

# Save file
ggsave(file= here("Graphics","histogram.png"),
       width = 4.5, height = 3,
       scale =2.2,
       dpi = 300)

```


## Check closest distance between any test point pair
```{r}
sp.test <- test_set
sp.test <- cbind(rownames(test_set), test_set)
df.test <- cbind(rownames(test_set), test_set)
colnames(sp.test) <- c('id', 'Easting', 'Northing', 'Load', 'Set')
coordinates(sp.test) <- ~Easting + Northing
class(sp.test)
# Calculate pairwise distances between points
d <- gDistance(sp.test, byid=T)
# Find second shortest distance (closest distance is of point to itself, therefore use second shortest)
min.d <- apply(d, 1, function(x) order(x, decreasing=F)[2])
new.test <- cbind(df.test, df.test[min.d,], apply(d, 1, function(x) sort(x, decreasing=F)[2]))
colnames(new.test) <- c('source',colnames(test_set), 'neighbor', 'Easting', 'Northing', 'Load', 'Set', 'distance')
# Return the nearest distance between any pair
dist_closestpair <- min(new.test$distance)
dist_closestpair
```

## Plot train and test points in Dataset 1
```{r}

ggplot(data = obs) + 
    geom_point(aes(x=Easting, y=Northing, size = Load, alpha = 0.2, shape = Set)) + 
    scale_color_manual(values = c("red", viridis::viridis(5)))+
    scale_shape_manual(values=c('Test'=19, 'Training'=1))+
    xlab("Easting") + ylab("Northing") + 
    theme(legend.position = "bottom") + 
    guides(alpha = FALSE, size = FALSE) +
    labs(subtitle = paste("DATASET 1: Distance between two closest test points = ", round(dist_closestpair, 2), sep = ""),
         x = "Easting (m)",
         y = "Northing (m)"
         #caption= paste("Fit to train set = ",round(best_fit, 2), sep = "")
    ) +
    theme_bw()

# Save file
ggsave(file= here("Graphics","train_test_ds1.png"),
       width = 4.5, height = 3,
       scale =2.2,
       dpi = 300)

```

## Add Dataset 2
```{r}


allpts_ds2$Set <- "Training"
allpts_ds2$Data <- "Dataset_2"
obs$Data <- "Dataset_1"
obs_ds1ds2 <- rbind(obs, allpts_ds2)

```

## Plot Dataset 1 and 2
```{r}

ggplot(data = obs_ds1ds2, aes(x=Easting, y=Northing, size = Load,color=Set, shape=Data, alpha=0.05)) + 
    geom_point() + 
    scale_color_manual(values = c("gray50", "red"))+
    scale_shape_manual(values=c('Dataset_1'=19, 'Dataset_2'=1))+
    xlab("Easting") + ylab("Northing") + 
    theme(legend.position = "bottom") + 
    guides(alpha = FALSE, size = FALSE) +
    labs(subtitle = paste("Distance between two closest test points = ", round(dist_closestpair, 2), sep = ""),
         x = "Easting (m)",
         y = "Northing (m)"
    ) +
    theme_bw()

# Save file
ggsave(file= here("Graphics","train_test_ds1ds2.png"),
       width = 4.5, height = 3,
       scale =2.2,
       dpi = 300)
```

## Export train and test point data seperately
```{r}

# Extract test and train rows
test <- subset(obs_ds1ds2, Set=="Test" )
train <- subset(obs_ds1ds2, Set=="Training" )

# Export test set to csv
write.csv(test, file = here("Train and Test", "test.csv"), row.names = FALSE)

```


## Export as InversionInput.txt; All weights = 1

Dataset 1 weight = 1
Dataset 2 weight = 1
Test points weight = 0

```{r}

# Format train points
train_input <- train[, c("Easting", "Northing", "Load")]
train_input$elevation <- 314.5 # Add elevation row
train_input <- train_input[, c("Easting", "Northing", "elevation", "Load")] # reorder columns 
train_input$Weight <- 1 # Add weight column
# Format test points
test_input <- test[, c("Easting", "Northing", "Load")]
test_input$elevation <- 314.5 # Add elevation row
test_input <- test_input[, c("Easting", "Northing", "elevation", "Load")] # reorder columns 
test_input$Weight <- 0 # Add weight column
# Combine train and test
input_all1 <- rbind(train_input, test_input)
write.table(input_all1, 
            file = here("Train and Test","inversionInput_all1.txt"), 
            sep = "\t",
            row.names = FALSE, col.names = FALSE)

```

## Export as InversionInput.txt; Weighted

Dataset 1 weight = 1
Dataset 2 weight = 0.5
Test points weight = 0

```{r}

# Extract dataset 1
train_input_ds1 <- subset(train, Data=="Dataset_1" )
train_input_ds1$Weight <- 1 # Add weight column
# Extract dataset 2
train_input_ds2 <- subset(train, Data=="Dataset_2" )
train_input_ds2$Weight <- 0.5 # Add weight column
# Combine
train_input_weighted <- rbind(train_input_ds1, train_input_ds2)
# Remove columns not needed for input text file
train_input_weighted <- train_input_weighted[, c("Easting", "Northing", "Load", "Weight")]
train_input_weighted$elevation <- 314.5 # Add elevation row
train_input_weighted <- train_input_weighted[, c("Easting", "Northing", "elevation", "Load", "Weight")] # reorder columns 
# Combine train and test
input_weighted <- rbind(train_input_weighted, test_input)

write.table(input_weighted,
            file = here("Train and Test","inversionInput_weighted.txt"),
            sep = "\t",
            row.names = FALSE, col.names = FALSE)

```


## Export as InversionInput.txt; No Dataset2

Dataset 1 weight = 1
Dataset 2 = removed
Test points weight = 0

```{r}

# Extract dataset 1
train_input_ds1 <- subset(train, Data=="Dataset_1" )
train_input_ds1$Weight <- 1 # Add weight column
train_input_ds1 <- train_input_ds1[, c("Easting", "Northing", "Load", "Weight")]
train_input_ds1$elevation <- 314.5 # Add elevation row
train_input_ds1 <- train_input_ds1[, c("Easting", "Northing", "elevation", "Load", "Weight")] # reorder columns 

# Combine train and test
input_nods2 <- rbind(train_input_ds1, test_input)

write.table(input_nods2,
            file = here("Train and Test","inversionInput_nods2.txt"),
            sep = "\t",
            row.names = FALSE, col.names = FALSE)

```