---
title: "Loss function selection"
author: "Michele Nguyen - edited by Maricar 12Oct2022"
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(goftest)
library(gridExtra)
library(moments)
library(LaplacesDemon)
```

## Checking loss function assumptions

Since each loss function is associated with distributional assumptions on the residuals, we can check the appropriateness of the choice of loss function based on the goodness-of-fit of the residuals to these assumed distributions.

### Mean squared error (MSE)

Using mean squared error as a loss function assumes a Gaussian distribution on the residuals. Other than compare the quantiles of the observed and assumed residual distributions, there are a few goodness-of-fit test statistics we can compute:

- Kolmogorov-Smirnov (K-S) test
- Cram'er-von Mises (CvM) test
- Anderson-Darling (AD) test
- Shapiro-Wilk test (for Gaussian distributions only)


```{r}

# Read in residual data (Dataset 1 fit only):

mse_resid <- read.csv("/Users/maricar/Github_repositories/tephra_v3.0/Runs/20220801_widerrange/Output CSV/s03_mse_wt_nods2_resid.csv")

head(mse_resid) # Here, Residual = Prediction - Load

```

```{r}

# Standardise residuals

mse_resid_sd <- sd(mse_resid$Residual)

mse_resid$SResidual <- mse_resid$Residual/sd(mse_resid$Residual)

```

```{r}

# Goodness of fit tests (typically also assume iid residuals):

# K-S test:
mse_ks <- ks.test(x = mse_resid$SResidual, y = pnorm)

# Anderson-Darling test:
mse_ad <- ad.test(x = mse_resid$SResidual, null = "pnorm")

# CvM test:
mse_cvm <- cvm.test(mse_resid$SResidual, null = "pnorm")

# Shapiro test:
mse_shapiro <- shapiro.test(mse_resid$SResidual)

```

```{r}

mse_skew <- skewness(mse_resid$Residual) 

```

```{r}

mse_qq <- ggplot(data = mse_resid, aes(sample = SResidual)) + stat_qq() + 
  geom_abline(slope = 1, intercept = 0) + theme_classic() + coord_equal(x = c(-4, 4), y = c(-4, 4)) + 
  labs(title="(A) MSE",
             x = "From model fit", 
              y = "Theoretical quantiles")+
    theme(plot.title=element_text(face="bold"))

```

### Chi-squared loss

```{r}

# Read in residual data (Dataset 1 fit only):

chi_resid <- read.csv("/Users/maricar/Github_repositories/tephra_v3.0/Runs/20220801_widerrange/Output CSV/s03_chi_wt_nods2_resid.csv")

head(chi_resid) # Here, Residual = Prediction - Load

```

```{r}

# Standardise residuals

# 1. Divide by square-root of Actual:
chi_resid$SResidual <- chi_resid$Residual/sqrt(chi_resid$Load)

# 2. Standardise by sd to get N(0,1):
chi_sresid_sd <- sd(chi_resid$SResidual)

chi_resid$SResidual <- chi_resid$SResidual/chi_sresid_sd

```

```{r}

# Goodness of fit tests (typically also assume iid residuals):

# K-S test:
chi_ks <- ks.test(x = chi_resid$SResidual, y = pnorm)

# Anderson-Darling test:
chi_ad <- ad.test(x = chi_resid$SResidual, null = "pnorm")

# CvM test:
chi_cvm <- cvm.test(chi_resid$SResidual, null = "pnorm")

# Shapiro test:
chi_shapiro <- shapiro.test(chi_resid$SResidual)

```

```{r}

chi_skew <- skewness(chi_resid$Residual) 

```

```{r}

chi_qq <- ggplot(data = chi_resid, aes(sample = SResidual)) + stat_qq() + 
  geom_abline(slope = 1, intercept = 0) + theme_classic() + coord_equal(x = c(-4, 4), y = c(-4, 4)) + 
  ggtitle("(B) Chi-square") + labs(x = "From model fit", 
                        y = "Theoretical quantiles")+
    theme(plot.title=element_text(face="bold"))

```

### Mean squared logarithmic error (MSLE)

```{r}

# Read in residual data (Dataset 1 fit only):

msl_resid <- read.csv("/Users/maricar/Github_repositories/tephra_v3.0/Runs/20220801_widerrange/Output CSV/s03_msl_wt_nods2_resid.csv")

head(msl_resid) # Here, Residual = Prediction - Load

```

```{r}

# Standardise residuals

# 1. Compute according to formula:
msl_resid$SResidual <- log10(msl_resid$Prediction + 1) - log10(msl_resid$Load + 1)

# 2. Standardise by sd to get N(0,1):
msl_sresid_sd <- sd(msl_resid$SResidual)

msl_resid$SResidual <- msl_resid$SResidual/msl_sresid_sd

```

```{r}

# Goodness of fit tests (typically also assume iid residuals):

# K-S test:
msl_ks <- ks.test(x = msl_resid$SResidual, y = pnorm)

# Anderson-Darling test:
msl_ad <- ad.test(x = msl_resid$SResidual, null = "pnorm")

# CvM test:
msl_cvm <- cvm.test(msl_resid$SResidual, null = "pnorm")

# Shapiro test:
msl_shapiro <- shapiro.test(msl_resid$SResidual)

```

```{r}

msl_skew <- skewness(msl_resid$Residual) 

```

```{r}

msl_qq <- ggplot(data = msl_resid, aes(sample = SResidual)) + stat_qq() + 
  geom_abline(slope = 1, intercept = 0) + theme_classic() + coord_equal(x = c(-4, 4), y = c(-4, 4)) + 
  ggtitle("(C) MSLE") + labs(x = "From model fit", 
                        y = "Theoretical quantiles")+
    theme(plot.title=element_text(face="bold"))

```

### Mean absolute error (MAE)

```{r}

# Read in residual data (Dataset 1 fit only):

mae_resid <- read.csv("/Users/maricar/Github_repositories/tephra_v3.0/Runs/20220801_widerrange/Output CSV/s03_mae_wt_nods2_resid.csv")

head(mae_resid) # Here, Residual = Prediction - Load

```

```{r}

# Standardise residuals

# 1. Compute according to formula:
mae_resid_scale <- mean(abs(mae_resid$Residual - median(mae_resid$Residual)))
#MLE estimator of scale = mean absolute deviation. (Fix mean to be 0.)

# 2. Standardise by scale to get Laplace(0, 1):
mae_resid$SResidual <- mae_resid$Residual/mae_resid_scale

```

```{r}

# Goodness of fit tests (typically also assume iid residuals):

# K-S test:
mae_ks <- ks.test(x = mae_resid$SResidual, y = plaplace)

# Anderson-Darling test:
mae_ad <- ad.test(x = mae_resid$SResidual, null = "plaplace")

# CvM test:
mae_cvm <- cvm.test(mae_resid$SResidual, null = "plaplace")

# Shapiro test:
mae_shapiro <- NA

```

```{r}

mae_skew <- skewness(mae_resid$Residual) 

```

```{r}

mae_qq <- ggplot(data = mae_resid, aes(sample = SResidual)) + 
  stat_qq(distribution = qlaplace) + 
  geom_abline(slope = 1, intercept = 0) + theme_classic() + 
  coord_equal(x = c(-7, 7), y = c(-7, 7)) + 
  ggtitle("(A) MAE") + labs(x = "From model fit", 
                        y = "Theoretical quantiles")+
    theme(plot.title=element_text(face="bold"))

```

### Mean absolute percentage error (MAPE)

```{r}

# Read in residual data (Dataset 1 fit only):

map_resid <- read.csv("/Users/maricar/Github_repositories/tephra_v3.0/Runs/20220801_widerrange/Output CSV/s03_map_wt_nods2_resid.csv")

head(map_resid) # Here, Residual = Prediction - Load

```




```{r}

# Standardise residuals

# 1. Compute according to formula:
map_resid$SResidual <- map_resid$Residual/map_resid$Load
map_sresid_scale <- mean(abs(map_resid$SResidual- median(mae_resid$SResidual))) 
#MLE estimator of scale = mean absolute deviation. (Fix mean to be 0.)

# 2. Standardise by scale to get Laplace(0, 1):
map_resid$SResidual <- map_resid$SResidual/map_sresid_scale

```

```{r}

# Goodness of fit tests (typically also assume iid residuals):

# K-S test:
map_ks <- ks.test(x = map_resid$SResidual, y = plaplace)

# Anderson-Darling test:
map_ad <- ad.test(x = map_resid$SResidual, null = "plaplace")

# CvM test:
map_cvm <- cvm.test(map_resid$SResidual, null = "plaplace")

# Shapiro test:
map_shapiro <- NA

```

```{r}

map_skew <- skewness(map_resid$Residual) 

```

```{r}

map_qq <- ggplot(data = map_resid, aes(sample = SResidual)) + 
  stat_qq(distribution = qlaplace) + 
 geom_abline(slope = 1, intercept = 0) + theme_classic() + 
  coord_equal(x = c(-7, 7), y = c(-7, 7)) + 
  ggtitle("(B) MAPE") + labs(x = "From model fit", 
                        y = "Theoretical quantiles")+
    theme(plot.title=element_text(face="bold"))

```


### Comparing loss functions

```{r}

temp_df <- data.frame("Loss function" = c("MSE", "Chi-squared", "MSLE",
                                          "MAE", "MAPE"), 
                      "K-S" = round(c(mse_ks$statistic, chi_ks$statistic, 
                                      msl_ks$statistic, mae_ks$statistic,
                                      map_ks$statistic), digits = 3),
                       "K-S p-value" = round(c(mse_ks$p.value, chi_ks$p.value, 
                                               msl_ks$p.value, mae_ks$p.value,
                                               map_ks$p.value), digits = 3),
                        "CvM" = round(c(mse_cvm$statistic, chi_cvm$statistic, 
                                        msl_cvm$statistic, mae_cvm$statistic,
                                        map_cvm$statistic), digits = 3),
                        "CvM p-value" = round(c(mse_cvm$p.value, chi_cvm$p.value, 
                                                msl_cvm$p.value, mae_cvm$p.value,
                                                map_cvm$p.value), digits = 3),
                        "AD" = round(c(mse_ad$statistic, chi_ad$statistic, 
                                       msl_ad$statistic, mae_ad$statistic,
                                       map_ad$statistic), digits = 3),
                        "AD p-value" = round(c(mse_ad$p.value, chi_ad$p.value, 
                                               msl_ad$p.value, mae_ad$p.value,
                                               map_ad$p.value), digits = 3),
                        "Shapiro-Wilk" = round(c(mse_shapiro$statistic, 
                                                 chi_shapiro$statistic, 
                                                 msl_shapiro$statistic,
                                                 mae_shapiro,
                                                 map_shapiro), 
                                               digits = 3),
                        "Shapiro-Wilk p-value" = round(c(mse_shapiro$p.value, 
                                                         chi_shapiro$p.value, 
                                                         msl_shapiro$p.value,
                                                         mae_shapiro,
                                                         map_shapiro
                                                         ),
                                                       digits = 3),
                        "Skewness" = round(c(mse_skew, chi_skew, msl_skew,
                                             mae_skew, map_skew),
                                                       digits = 3)
                         )

knitr::kable(temp_df, caption = "Goodness-of-fit test statistics and p-values for the model residuals according to the loss functions used to fit them. The estimated skewness of the raw residuals (Predicted - Actual) are also shown. For all stests, the null hypothesis is that the residuals follow the distribution assumed by the choice of the loss function. So large p-values indicate better fit and adherence to the assumptions.")

```

The skewness estimates are not a good reflection of the tendency for symmetrric loss functions to overpredict rather than underpredict (since the load values themselves cannot physically go below zero). Instead, we visualise this using density plots of the raw residuals (Predicted - Actual) against Actual.

```{r}

den_mse <- density(mse_resid$Residual, from = -60, to = 60, n = 512)
# plot(den_mse$x, den_mse$y)
den_chi <- density(chi_resid$Residual, from = -60, to = 60, n = 512)
den_msl  <- density(msl_resid$Residual, from = -60, to = 60, n = 512)

den_mae  <- density(mae_resid$Residual, from = -60, to = 60, n = 512)
den_map  <- density(map_resid$Residual, from = -60, to = 60, n = 512)

den_df <- data.frame("Residual" = rep(den_mse$x, 5), 
                     "Density" = c(den_mse$y, den_chi$y, den_msl$y, den_mae$y, den_map$y),
                     "Loss" = rep(c("MSE", "Chi-squared", "MSLE", "MAE", "MAPE"), each = length(den_mse$x)))

```

```{r}
den_compare<- ggplot() + geom_line(data = den_df, aes(x = Residual, y = Density, color = Loss)) + 
  theme_classic() + ggtitle("Density of raw residuals") + coord_cartesian(xlim = c(-20, 20)) +
  geom_vline(xintercept = 0, lty = 2)

den_compare

```

```{r}

# Residual vs load by color:

rvsl_df <- data.frame("Residual" = c(mse_resid$Residual, chi_resid$Residual, msl_resid$Residual,
                                     mae_resid$Residual, map_resid$Residual),
                      "Load" = c(mse_resid$Load, chi_resid$Load, msl_resid$Load,
                                     mae_resid$Load, map_resid$Load),
                     "Loss" = rep(c("MSE", "Chi-squared", "MSLE", "MAE", "MAPE"), each = nrow(mse_resid)))

```

```{r}

rvsl_compare <- ggplot() + geom_point(data = rvsl_df, aes(x = Residual, y = Load, color = Loss)) + 
  theme_classic() + ggtitle("Raw residuals against load") 

rvsl_compare

```


```{r, fig.height=3.5, fig.width=10.5}

rvsl_mse <- ggplot() + geom_point(data = mse_resid, aes(x = Residual, y = Load)) + coord_fixed(xlim = c(-100,100), ylim = c(0, 175)) + theme_classic() + ggtitle("MSE") + geom_abline(intercept = 25, slope = 0, lty = 2) + geom_hline(yintercept = 25, lty = 2) + geom_vline(xintercept = 25, lty = 2)

rvsl_chi <- ggplot() + geom_point(data = chi_resid, aes(x = Residual, y = Load))  + coord_fixed(xlim = c(-100,100), ylim = c(0, 175)) + theme_classic() + ggtitle("Chi-squared") + geom_hline(yintercept = 25, lty = 2) + geom_vline(xintercept = 25, lty = 2)

rvsl_msl <- ggplot() + geom_point(data = msl_resid, aes(x = Residual, y = Load))  + coord_fixed(xlim = c(-100,100), ylim = c(0, 175)) + theme_classic() + ggtitle("MSLE") + geom_hline(yintercept = 25, lty = 2) + geom_vline(xintercept = 25, lty = 2)

grid.arrange(rvsl_mse, rvsl_chi, rvsl_msl, nrow = 1, ncol = 3)

```

```{r, fig.height=3.5, fig.width=7}

rvsl_mae <- ggplot() + geom_point(data = mae_resid, aes(x = Residual, y = Load)) + coord_fixed(xlim = c(-100,100), ylim = c(0, 175)) + theme_classic() + ggtitle("MAE") + geom_abline(intercept = 25, slope = 0, lty = 2) + geom_hline(yintercept = 25, lty = 2) + geom_vline(xintercept = 25, lty = 2)

rvsl_mape <- ggplot() + geom_point(data = map_resid, aes(x = Residual, y = Load))  + coord_fixed(xlim = c(-100,100), ylim = c(0, 175)) + theme_classic() + ggtitle("MAPE") + geom_hline(yintercept = 25, lty = 2) + geom_vline(xintercept = 25, lty = 2)

grid.arrange(rvsl_mae, rvsl_mape, nrow = 1, ncol = 2)

```

```{r, fig.height=3.5, fig.width=10.5}

qq1 <- grid.arrange(mse_qq, chi_qq, msl_qq,
             nrow = 1, ncol = 3)

ggsave(qq1, file= here("qq1.png"),
       #width = 7, height = 3.5,
       scale =1,
       dpi = 300)

```

```{r, fig.height=3.5, fig.width=7}

qq2 <- grid.arrange(mae_qq, map_qq, nrow = 1, ncol = 2)

ggsave(qq2, file= here("qq2.png"),
       #width = 7, height = 3.5,
       scale =1,
       dpi = 300)

```